### 2019-02-27 sections/intuition.tex
spelling
use more concrete examples in the intuition section
concentrate on the intuition: introduce symbols for tasks only, 
everthing else is introduced in the syntax section, less meta variables, 
no refereces to semantic functions

@@

The following program shows the different collaboration operators in action.
Users have a choice ($\Xor$) whether they want tea or coffee.
They always get an egg.
The drink and the food are prepared in parallel ($\And$).
When drink and food are prepared, users can step ($\Next$) to eating the results.

@@

The way the combinators are defined matches real life closely.
When we want to have breakfast, we have to complete several other tasks first before we can do so.
We decide what we want to have and then prepare it.
We can prepare the different items we have for breakfast in parallel, but not at the same time.
For example, it is impossible to scramble eggs, and put on the kettle for tea simultaneously.
Instead, what is meant by parallel is that the order in which we do tasks and the smaller tasks that they are composed of, does not matter.
\todo{Alejandro has placed a line and exclamation under this sentence. I'm not sure what to do here.}
Then finally, only when every item we want to have for breakfast is ready, can we sit down and enjoy it.
\end{example}

@@

Tasks are modular in the following ways.
First, larger tasks are composed from smaller ones.
Second, tasks are first-class, they can be arguments and results of functions.
Third, tasks can be values of other tasks.
These aspects make it possible for programmers to model custom collaboration patterns.

@@

In \TOPHAT, editors are the basic method of communication with the environment.
There are different editors, denoted by different box symbols.
An editor of the form $\Edit v$, where $v$ is of type $\tau$, reacts to two kinds of events:
Change events $v$ with a value of type $\tau$ and clear events $\Empty$ which empty the editor.
This results in a typed empty editor $\Enter \tau$, which is annotated with the same type.
\todo{Alejandro: unclear} %Nico: He has marked the above paragraph as being unclear.

The sole purpose of editors is to interact with users by remembering the last value that has been sent to them.
There are no output events.
As values of editors can be observed, for example by a user interface, editors serve as facilities for both input and output.
An empty editor $\Enter \tau$ stands for a prompt to input data, while a filled editor can be seen either as outputting a value, or as an input that comes with a default value.

@@

\todo{Alejandro: What is a value? What is a type?}

@@

Several observations can be made about tasks.
One of those is determining the value of a task $\Value(t)$.
%NOTE: Should we change the type of Value to Maybe v? Both johan and alejandro mention that it is not completely clear to them. --NN
%NOTE: I don't think that we should use Maybe. We should use the term partial function and bottom. --MKL
Not all tasks have a value, for example the empty editor $\Enter \tau$, which makes $\Value$ a partial function.
For example, $\Value(\Edit 5) = 5$, $\Value(\Enter \Int) = \bot$.

Another observation is the set of events a task can react to.
This is important for generating user interfaces.
For example, the task $\Edit 5$ can react to value events $v$ and empty events $\Empty$.

In order to render a user interface, the system needs to observe a task's user interface.
This is done compositionally.
User interfaces of combined tasks are composed of the user interfaces of the components.
For example of two tasks combined with a step combinator, only the left hand side is rendered.
Two parallel tasks are rendered next to each other.

In order for the system to actually render a task, it needs to observe the task's value, possible inputs, and user interface.
With this information, the system can then display the current state of the task, together with buttons that show the actions a user can engage in.

The final observation is to determine whether a task results in $\Fail$ (failure).
The step combinator $\Next$ and the choice combinator $\Xor$ use this to prevent users from picking a failing task.

@@

In a step $t \Next e$, the decision to move on is taken by $\Next$, not by $t$.
The decision is based on a speculative evaluation of $e$.
The step combinators in $t \Then e$ and $t \Next e$ pass the value $v$ of $t$ to the continuation $e$.
Both steps act like $t$ as long as the step is guarded.
A step is guarded if either the left task has no value, or the speculative evaluation $e\ v$ yields $\Fail$.
Once it becomes unguarded, the step continues as $e\ v$.
The task $t \Next e$ additionally requires a continue event $\Continue$ to proceed.
Speculative evaluation is designed so that possible side effects are undone.

The step combinators give rise to a form of internal communication:
They represent data flow that \emph{follows} control flow.

@@

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
Users can not directly interact with shared data, a shared editor is required for that.
If $r$ is a reference of type $\tau$, then $\Update r$ is an editor whose value is that of $r$.

@@

Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
This operator introduces interleaving concurrency.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ by $\First$ (first) and $\Second$ (second) respectively.
This unambiguously renames the inputs, removing any source of nondeterminism.

@@

\Citet{books/Hoare85CSP} states that nondeterminism is only ever useful for \emph{specifying} systems, never for implementing them.
\TOPHAT is meant solely for implementation and does not have any form of nondeterminism.
Input events for parallel tasks are disambiguated, internal steps have a well-defined evaluation order, and internal choice is left-biased.

>>>>

\subsection{Tasks model collaboration}
%NOTE: Heb toch maar deze kop toegevoegd, onderstaande gaat er duidelijk over, en de sprong van het voorbeeld naar "Tasks are reusable" is anders abrupt!

@@

The following program shows the different collaboration operators in the setting of making a breakfast.
Users have a choice ($\Xor$) whether they want tea or coffee.
They always get an egg.
The drink and the food are prepared in parallel ($\And$).
When drink and food are prepared, users can step ($\Next$) to eating the results.

@@

The way the combinators are defined matches real life closely.
When we want to have breakfast, we have to complete several other tasks first before we can do so.
We decide what we want to have and then prepare it.
We can prepare the different items we have for breakfast in parallel, but not at the same time.
For example, it is impossible to scramble eggs, and put on the kettle for tea simultaneously.
Instead, what is meant by parallel is that \emph{the order in which we do tasks and the smaller tasks that they are composed of, does not matter}.
%XXX: Alejandro has placed a line and exclamation under this sentence. I'm not sure what to do here. --NN
% I think he thinks it is an essential part of understanding tasks. I emphasised it. --TS
Then finally, only when every item we want to have for breakfast is ready, can we sit down and enjoy it.
\end{example}

@@

There are three ways in which tasks are modular.
First, larger tasks are composed from smaller ones.
Second, tasks are first-class, they can be arguments and results of functions.
Third, tasks can be values of other tasks.
These aspects make it possible for programmers to model custom collaboration patterns.

@@

In \TOPHAT, \emph{editors} are the basic method of communication with the environment.
Editors are modelled after input widgets from graphical user interfaces.
There are different editors, denoted by different box symbols.
Take for example an editor holding the integer seven: $\Edit 7$.
Such an editor reacts to two kinds of events.
The first are change events, for example the values $42$ or $37$, which are of the same type.
The second is a clear event $\Empty$, which empties the editor.
This results is an empty editor $\Enter \Int$, annotated to only hold integers.

%XXX: Alejandro: unclear
%XXX: Alejandro: What is a value? What is a type?
% He has marked the above paragraph as being unclear. --NN
% Vraag is alleen wat dan precies... :-( Ik heb er een concreet voorbeeld van gemaakt met integers. --TS

The sole purpose of editors is to interact with users by remembering the last value that has been sent to them.
There are no output events.
As values of editors can be observed, for example by a user interface, editors serve as facilities for both input and output.
An empty editor ($\Enter$) stands for a prompt to input data,
while a filled editor ($\Edit$) can be seen either as outputting a value, or as an input that comes with a default value.

@@

@@

Several observations can be made on tasks.
One of those is determining the value of a task.
%NOTE: Should we change the type of Value to Maybe v? Both Johan and Alejandro mention that it is not completely clear to them. --NN
% I don't think that we should use Maybe. We should use the term partial function and bottom. --MKL
% I agree! I made the examples here also concrete instead of introducing the value function. --TS
Not all tasks have a value, the empty editor for instance, which makes value observation partial.
For example, the value of $\Edit 7$ is $7$, but the value of $\Enter \Int$ is $\bot$.

Another observation is the set of input events a task can react to.
% This is important for generating user interfaces.
For example, the task $\Edit 7$ can react to value events and empty events, as discussed before.

In order to render a task, we need to observe a task's user interface.
This is done compositionally.
User interfaces of combined tasks are composed of the user interfaces of the components.
For example, of two tasks combined with a step combinator, only the left hand side is rendered.
Two parallel tasks are rendered next to each other.
Combining this information with the task's value and possible inputs,
we can display the current state of the task,
together with buttons that show the actions a user can engage in.

The final observation is to determine whether a task results in a failure, denoted by $\Fail$.
The step combinator $\Next$ and the choice combinator $\Xor$ use this to prevent users from picking a failing task.

@@

In a step $t \Next e$, the decision to move on from a task $t$ to its continuation $e$ is taken by $\Next$, not by~$t$.
The decision is based on a speculative evaluation of $e$.
The step combinator in $t \Next e$ passes the value $v$ of $t$ to the continuation $e$.
Steps act like $t$ as long as the step is guarded.
A step is guarded if either the left task has no value, or the speculative evaluation of $e$ applied to $v$ yields the failure task $\Fail$.
Once it becomes unguarded, the step continues as the result of $e\ v$.
Speculative evaluation is designed so that possible side effects are undone.
The task $t \Next e$ additionally requires a continue event $\Continue$ to proceed.
% Its cousin $\Then$ makes a step without user interaction.

Step combinators give rise to a form of internal communication.
They represent data flow that \emph{follows} control flow.

@@

The step combinator is one form of internal communication, where task values are passed to continuations.
Another form of internal communication is shared data.
Shared data enables data flow \emph{across} control flow, in particular between parallel tasks.
Shared data sources are assignable references whose changes are immediately visible to all tasks interested in them.
Users can not directly interact with shared data, a shared editor is required for that.
If $x$ is a reference of type $\tau$, then $\Update x$ is an editor whose value is that of $x$.

@@

Let $t_1$ and $t_2$ be tasks.
The parallel combination $t_1 \And t_2$ stands for two independent tasks carried out at the same time.
This operator introduces interleaving concurrency.
For the system it does not matter if the tasks are executed by two people actually in parallel, or by one person who switches between the tasks.
The inputs sent to the component tasks are interleaved into a serial stream, which is sent to the parallel combinator.
We assume that such a serialization is always possible.
The tasks are truly independent of each other, all interleavings are permitted.
The environment must prefix events to $t_1$ and $t_2$ respectively by $\First$ (first) and $\Second$ (second).
This unambiguously renames the inputs, removing any source of nondeterminism.

@@

\Citet{books/Hoare85CSP} states that nondeterminism is only ever useful for \emph{specifying} systems, never for implementing them.
\TOPHAT is meant solely for implementation and does not have any form of nondeterminism.
Input events for parallel tasks are disambiguated, internal steps ($\Then$) have a well-defined evaluation order, and internal choice ($\Or$) is left-biased.
